Automated Lensing Learner - I: An automated Strong Lensing Identification Pipeline

### 2017-03-09 ms.tex
Finished subsection on training set size and model performance/model train time

%-------------------------------------------------%
\begin{abstract} 
Gravitational lensing offers a direct probe of the underlying mass
distribution of lensing systems, a window to the high redshift
universe, and a geometric probe of cosmological models.  The advent of
large scale surveys such as the Large Synoptic Sky Telescope and
Euclid has prompted a need for automatic and efficient identification
of strong lensing systems.  We present {\em (ALL) Automated Lensing
  Learner}, a strong lensing identification pipeline that will be
publicly released as open source software.  In this first application,
we employ a fast feature extraction method, Histogram of Oriented
Gradients (HOG), to capture edge patterns that are characteristic of
strong gravitational arcs in galaxy-galaxy lensing.  We use logistic
regression to train a supervised classifier model on the HOG of HST-
and LSST-like images.  We use an area under the curve (AUC) of a
Receiver Operating Characteristic curve to assess model performance;
$AUC=1.0$ is an ideal trained classifier, and $AUC=0.5$ does not
provide any classification information.  Our best performing models
exhibit an AUC of \todo{XX} for an HST-like sample.  However, for one
orbit of LSST, our model only reaches an AUC of \todo{XX}.  In absence
of the lens galaxy in one orbit of LSST-like data, the AUC improves to
\todo{XX}.  For 10-year mock LSST observations, the respective AUC
with and without the lens galaxy improves to \todo{XX} and \todo{XX}.
Our results show efficient and effective methods for automated strong
lensing identification with \wording{scalable linear methods that are
  easy to parallelize with existing open source tools.}
\end{abstract}
%-------------------------------------------------%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                                        
\begin{figure}[t]\label{fig:ROCcompilation}
\begin{center}
\includegraphics[width=\columnwidth]{figures/metrics/ROC/ROC_compilation.pdf}
\caption{Red, blue, green: ROC curves for models trained on our whole
  10,000 training set and tested on our holdout set of 1,000.  These
  respectively correspond to the HST, LSST 1 year, and LSST 10 year
  data.  The solid lines are for data that include the lensing central
  galaxy, and the dashed lines for the data where there is no lensing
  galaxy, mimicking an ideal removal of the lens.  Model performance
  can be summarized by the area under the curve (AUC), labeled in the
  legend. $AUC=1$ is a perfect model, and $AUC=0.5$ is a useless
  model.}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%            
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                                        
\begin{figure}[t]\label{fig:PRcompilation}
\begin{center}
\includegraphics[width=\columnwidth]{figures/metrics/PR/PR_compilation.pdf}
\caption{Red, blue, green: Precision-recall (PR) curves for models
  trained on our whole 10,000 training set and tested on our holdout
  set of 1,000.  These respectively correspond to the same models and
  data shown in Figure~\ref{fig:ROCcompilation}.  An ideal model would
  reach both a precision (purity) and recall (completeness) that equal
  1.  Note, that this performance describes a data set with a 50-50
  split between lens and non-lens containing images.}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%-------------------------------------------------%
\subsection{Effects of Dataset Size on Model Performance}\label{sec:datasize}
%-------------------------------------------------% 
%-------------------------------------------------% 
\subsubsection{Effects of Training Set Size}\label{sec:trainsetsize}
%-------------------------------------------------% 
In this section, we show the effects of training set size on model
performance.  Figure~\ref{fig:trainsizeLSST10} shows how the AUC and
the logistic regression coefficient depend on the log of the size of
the training set.  In solid blue lines, we see that the AUC improves
almost linearly with the log of the training size set.  In dotted red,
we see that the train time roughly scales logarithmically with the log
of the train size, with an exception of the data point corresponding
to train size of 2000.

Each model was trained with the same HOG parameterization from the
grid search and the optimal regularization parameter for that subset
train data.  The train time of a given model generally increases for
increasing regularization parameter.  For the subset of train size
2000, the optimal regularization parameter happened to be
$C_{LogReg}=100$, whereas it was $C_{LogReg}=500$ for the subset of trainsize
1000, and $C_{LogReg}=1000$ for the subset of trainsize 4000.

% Scores vs. regularization on larger train/test set with comparable
% grid search in parameterization
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                                        
\begin{figure}[t]
\begin{center}
\includegraphics[width=\columnwidth]{figures/trainsize/LSST10_with_nLSST10.pdf}
\caption{Solid (dotted) blue line: AUC for models with varying size of
  the training set for LSST10 (nLSST10).  Solid (dotted) red line:
  Optimal regularization parameter for training on the train size for
  LSST10 (nLSST10). The improvement of AUC scales linearly with the
  log of the training set size, but train time roughly scales
  logarithmically.  Note, LSST10 requires more model complexity to
  exceed the performance of nLSST10.}\label{fig:trainsizeLSST10}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%-------------------------------------------------% 
\subsubsection{Effects of Rotation on AUC}
%-------------------------------------------------% 
To augment our training set, we rotated each image in the set by
multiples of $90^o$.  The augmentation (factor of four) naturally
optimizes the use of available training data, since our feature
extraction method of HOG is not rotationally invariant.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                                        
\begin{figure}[t]\label{fig:rotation_test}
\begin{center}
\includegraphics[width=\columnwidth]{figures/rotated_aucs/rotated_aucs.pdf}
\caption{The AUCs of models trained on the full 10,000 and tested on
  the holdout 1,000.  Blue circles: AUC calculated from scores of
  images at a given rotation (e.g. 0, 90, 180, and 270 degrees).  Red
  stars: AUC calculated from the average score of all rotations of
  each image.  The average score produces an improved AUC in all data
  sets.}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                                        
\begin{figure*}[t]\label{fig:ROCsamplesLSST10}
\begin{center}
\includegraphics[width=.8\columnwidth]{figures/tiles/compilations/LSST10/tp.pdf}\hspace{5pt}
\includegraphics[width=.8\columnwidth]{figures/tiles/compilations/LSST10/fp.pdf}\\\vspace{5pt}
\includegraphics[width=.8\columnwidth]{figures/tiles/compilations/LSST10/bp.pdf}\hspace{5pt}
\includegraphics[width=.8\columnwidth]{figures/tiles/compilations/LSST10/bn.pdf}\\\vspace{5pt}
\includegraphics[width=.8\columnwidth]{figures/tiles/compilations/LSST10/fn.pdf}\hspace{5pt}
\includegraphics[width=.8\columnwidth]{figures/tiles/compilations/LSST10/tn.pdf}\\
\caption{LSST mock images over 10 years. Left two columns: Lens
  containing images, annotated with the image score assigned by our
  trained classifier.  Right two columns: Non-lens containing images,
  annotated with the image score.  The top two rows show
  characteristic images that will be accepted with a high threshold
  for classification, contributing to the bottom left of the ROC
  curve.  The middle two rows show characteristic images that will be
  accepted with a moderate threshold, contributing to the knee of the
  ROC curve.  The bottom two rows show characteristic images that will
  only be accepted with an extremely lenient threshold, contributing
  to the top right area of the ROC curve.}
\end{center}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

%-------------------------------------------------%
\begin{abstract} 
Gravitational lensing offers a direct probe of the underlying mass
distribution of lensing systems, a window to the high redshift
universe, and a geometric probe of cosmological models.  The advent of
large scale surveys such as the Large Synoptic Sky Telescope and
Euclid has prompted a need for automatic and efficient identification
of strong lensing systems.  We present {\em (ALL) Automated Lensing
  Learner}, a strong lensing identification pipeline that will be
publicly released as open source software.  In this first application,
we employ a fast feature extraction method, Histogram of Oriented
Gradients (HOG), to capture edge patterns that are characteristic of
strong gravitational arcs in galaxy-galaxy lensing.  We use logistic
regression to train a supervised classifier model on the HOG of HST-
and LSST-like images.  We use an area under the curve (AUC) of a
Receiver Operating Characteristic curve to assess model performance;
$AUC=1.0$ is an ideal trained classifier, and $AUC=0.5$ does not
provide any classification information.  Our best performing models on
a training set of 10,000 lens containing images and 10,000 non-lens
containing images exhibit an AUC of 0.975 for an HST-like sample.
However, for one orbit of LSST, our model only reaches an AUC of
0.625.  For 10-year mock LSST observations, the AUC improves to 0.809.
Model performance appears to continually improve with the size of the
training set.  Models trained on fewer images perform better in
absence of the lens galaxy, which can decrease the false positive rate
by blending with light from a lensed galaxy image near the central
lensing galaxy.  However, with larger training data sets, information
from the lensing galaxy improves model performance.  \wordchoice{Our
  results show efficient and effective methods for automated strong
  lensing identification with scalable linear methods that are easy to
  parallelize with existing open source tools and are reproducible on
  personal laptop computers.}
\end{abstract}
%-------------------------------------------------%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                                        
\begin{figure}[t]
\begin{center}
\includegraphics[width=\columnwidth]{figures/metrics/ROC/ROC_compilation.pdf}
\caption{Red, blue, green: ROC curves for models trained on our whole
  10,000 training set and tested on our holdout set of 1,000.  These
  respectively correspond to the HST, LSST 1 year, and LSST 10 year
  data.  The solid lines are for data that include the lensing central
  galaxy, and the dashed lines for the data where there is no lensing
  galaxy, mimicking an ideal removal of the lens.  Model performance
  can be summarized by the area under the curve (AUC), labeled in the
  legend. $AUC=1$ is a perfect model, and $AUC=0.5$ is a useless
  model.}\label{fig:ROCcompilation}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%            
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                                        
\begin{figure}[t]
\begin{center}
\includegraphics[width=\columnwidth]{figures/metrics/PR/PR_compilation.pdf}
\caption{Red, blue, green: Precision-recall (PR) curves for models
  trained on our whole 10,000 training set and tested on our holdout
  set of 1,000.  These respectively correspond to the same models and
  data shown in Figure~\ref{fig:ROCcompilation}.  An ideal model would
  reach both a precision (purity) and recall (completeness) that equal
  1.  Note, that this performance describes a data set with a 50-50
  split between lens and non-lens containing
  images.}\label{fig:PRcompilation}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%-------------------------------------------------%
\subsection{Effects of Dataset Size on Model Performance}\label{sec:datasize}
%-------------------------------------------------% 
%-------------------------------------------------% 
\subsubsection{Effects of Training Set Size}\label{sec:trainsetsize}
%-------------------------------------------------%

In this section, we show the effects of training set size on model
performance on the holdout test set of 1,000 images.  We also compare
the improvement between images that include the lens galaxy and images
with no lens galaxy.

Figure~\ref{fig:trainsizeLSST10} shows how the AUC and the optimal
logistic regression coefficient depend on the log of the size of the
training set for both the LSST10 data in solid blue lines, and the
nLSST10 data in dashed blue lines.  The AUC for models trained on the
LSST10 data improves almost linearly with the log of the training set
size, increasing from AUC=0.705 to AUC=0.788 when the train size is
increased from 500 lens/non-lens images to 8000 lens/non-lens images.
However, for the nLSST10 data, where the lens galaxy has been removed
from the images, the improvement is less dramatic.  With the same
increase in training size, the AUC for nLSST10 changes from just below
0.77 to just below 0.78.

In solid red, we see that the optimal value for $C_\text{LogReg}$ for
LSST10 roughly scales logarithmically with the log of the train size,
with an exception of the data point corresponding to train size of
2000.  Generally, a larger training set will require an increase in
model complexity to capture additional information.  This is also true
for the optimal $C_\text{LogReg}$ dependence on the number of training
images in the nLSST10 data.  But, the required complexity is
systematically less than for the LSST10 images.

% Comparison of the two LSST10 vs. nLSST10
In the LSST10 case, the trained model can incorporate the additional
information of the edges from the lens galaxy, which is correlated
with the lensing cross-section and likelihood of the image being a
lensing system.  The nLSST10 images do not contain this information,
but provide cleaner signals of the lensed image for lensed images that
occur close to the lens galaxy.  The cleaner signal in nLSST10 allows
for better model performance for smaller training data set sizes
($N_\text{train}\sim5,000$).  However, models trained on LSST10
improve more rapidly with train size, since the additional information
from the lens galaxy better describes the lens containing images.  For
$N_\text{train}\gsim 5,000$, the models trained on LSST10 outperform
  those trained on nLSST10.

%M Model complexity, train time, and performance
Since the models that contain information from the lens galaxy edges
in LSST10 are more complex, the models require a larger training set
size for a better fit.  While model performance for LSST10 appears to
steadily increase, this comes at the cost of increased train time,
which is two-fold.  The train time will increase due to both an
increase in data to fit, and also an increase in optimal
$C_\text{LogReg}$ where the volume of hyperparameter space for allowed
solutions is larger (see red lines in
Figure~\ref{fig:regularization}).

% Train time
Figure~\ref{fig:traintimeLSST10} shows the train time for LSST10 and
nLSST10 as a function of the size of the training set for the optimal
regularization parameter for that subset training data.  Each model
uses features extracted with the same HOG parameterization from the
grid search and the optimal regularization parameter for that subset
train data.  The train time of a given model generally increases for
increasing regularization parameter.  For the subset of train size
$N_\text{train}=2000$ in LSST10, the optimal regularization parameter
happened to be $C_{LogReg}=100$, whereas it was $C_{LogReg}=500$ for
the subset of trainsize 1000, and $C_{LogReg}=1000$ for the subset of
trainsize 4000.  This makes the train time increase at train size
$N_\text{train}=2000$ for LSST10 less dramatic than the average
log-log slope of approximately 2.

% Train time behavior for nLSST10
Since nLSST10 does not contain the lens galaxies, fewer of the
extracted features describe the lensing system, requiring decreased
model complexity.  The increase in train time for nLSST10 as a
function of train size is mostly due to only having more data to fit
in the regression, leading to a steady and slow increase of train time
with number of training images with log-log slope of approximately 1.

% AUC vs. regularization on larger train/test set with comparable
% grid search in parameterization
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                                        
\begin{figure}[t]
\begin{center}
\includegraphics[width=\columnwidth]{figures/trainsize/LSST10_with_nLSST10.pdf}
\caption{Solid (dotted) blue line: AUC for models with varying size of
  the training set for LSST10 (nLSST10).  Solid (dotted) red line:
  Optimal regularization parameter for training on the train size for
  LSST10 (nLSST10). The improvement of AUC scales linearly with the
  log of the training set size, but train time roughly scales
  logarithmically.  Note, LSST10 requires more model complexity to
  exceed the performance of nLSST10.}\label{fig:trainsizeLSST10}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                                        
\begin{figure}[t]
\begin{center}
\includegraphics[width=\columnwidth]{figures/traintime/LSST10.pdf}
\caption{Solid (dotted) blue line: Train time for models with varying
  size of the training set for LSST10 (nLSST10).  Train time roughly
  scales logarithmically.  Note, LSST10 requires more model complexity
  to exceed the performance of nLSST10, and therefore requires more
  training time for continual increase in
  performance.}\label{fig:traintimeLSST10}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%-------------------------------------------------% 
\subsubsection{Effects of Rotation on AUC}
%-------------------------------------------------% 
To augment our training set, we rotated each image in the set by
multiples of $90^o$.  Augmenting our data by a factor of four
naturally optimizes the use of available training data, since our
feature extraction method of HOG is not rotationally invariant.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                                        
\begin{figure}[t]\label{fig:rotation_test}
\begin{center}
\includegraphics[width=\columnwidth]{figures/rotated_aucs/rotated_aucs.pdf}
\caption{Our summary figure: The AUCs of models trained on the full
  10,000 and tested on the holdout 1,000.  Blue circles: AUC
  calculated from scores of images at a given rotation (e.g. 0, 90,
  180, and 270 degrees).  Red stars: AUC calculated from the average
  score of all rotations of each image.  The average score produces an
  improved AUC in all data sets.}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                                        
\begin{figure*}[t]\label{fig:ROCsamplesLSST10}
\begin{center}
\includegraphics[width=.8\columnwidth]{figures/tiles/compilations/LSST10/tp.pdf}\hspace{5pt}
\includegraphics[width=.8\columnwidth]{figures/tiles/compilations/LSST10/fp.pdf}\\\vspace{5pt}
\includegraphics[width=.8\columnwidth]{figures/tiles/compilations/LSST10/bp.pdf}\hspace{5pt}
\includegraphics[width=.8\columnwidth]{figures/tiles/compilations/LSST10/bn.pdf}\\\vspace{5pt}
\includegraphics[width=.8\columnwidth]{figures/tiles/compilations/LSST10/fn.pdf}\hspace{5pt}
\includegraphics[width=.8\columnwidth]{figures/tiles/compilations/LSST10/tn.pdf}\\
\caption{LSST 10 year mock images. Left two columns: Lens containing
  images, annotated with the image score assigned by our trained
  classifier.  Right two columns: Non-lens containing images,
  annotated with the image score.  The top two rows show
  characteristic images that will be accepted with a high threshold
  for classification, contributing to the bottom left of the ROC curve
  in Figure~\ref{fig:ROCcompilation}.  The middle two rows show
  characteristic images that will be accepted with a moderate
  threshold, contributing to the knee of the ROC curve, with true
  positive rates and false positive rates of $tpr\approx0.8$ and
  $fpr\approx0.2$, respectively.  The bottom two rows show
  characteristic images that will only be accepted with an extremely
  lenient threshold, contributing to the top right area of the ROC
  curve.}
\end{center}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
